Interactive Graphics for Visually Diagnosing Forest Classifiers in R

### 2022-06-29 response_reviewers.Rmd

Springer Journals Editorial Office
%JOURNALFULLTITLE

\textcolor{violet}{Thank you. We appreciate the chance to do a further revision, and hope that these changes satisfy the AE and reviewer.}

## Response to Reviewer 2

\textcolor{violet}{We appreciate all of your feedback which help to improve our paper. 
 }

@@

Summary of the paper

The paper introduces an app for visual exploration of tree-based ensemble models in three steps. First, the app enables the user to investigate how well the model performs when it comes to predicting each class in its observation-level functionality. Second, the tree-level section allows the user to select a tree based on a variable importance measure in one of the first three nodes and visualize the way the first three splits are made in that tree. Finally, the app visually compares the predictive performance of the model to a benchmark one, which is illustrated by comparing a projection-pursuit-based random forest to a standard one.

The paper and its accompanying app seem to be a good fit for Computational Statistics, as they apply interesting visualization techniques to a tree-based ensemble model that will surely prove useful for practitioners.

However, for this potential to materialize the authors should put more emphasis on the limitations of their approach. In particular, the app focuses on the first three nodes of the trees in question, which would be much less informative in settings requiring deep trees. \textcolor{violet}{The app was modified to adress this issue. In the new application the user can select specific nodes to see (from one to all the nodes). If there are a lot of nodes will be better to select 4 or less each time to see better the complete display but the application is not limited to see just the first three nodes.}

@@

Furthermore, plots such as the parallel coordinate plot would lose their appeal in settings with thousands of variables. A brief discussion of circumstances limiting the usability of the app would address these concerns and allow users to easily determine whether the app is suitable for their application.

\textcolor{violet}{In the discussion section we incorporate the app limitations.}

@@

\item p. 11, line 16: As section 3 discusses all other plots produced by the app it would be useful if it also included an overview of the parallel coordinate plot that only appears in section 4, so the meaning of that plot is not immediately clear (e.g. what is displayed on the vertical axis).

\textcolor{violet}{We have included a description of parallel coordinate plot in Section 4 were this figure appears and the corresponding reference for further details. In the vertical axis the values for each variable are standardized to easily compare them.}

\item p. 13, line 38: I find the choice of the interaction driver for the model tab puzzling. I would argue that the key statistic distinguishing the trees in the ensemble is OOB error so a more natural design choice would be to include jittered points in the OOB error boxplot so that the user can explore the well or poorly performing trees in more detail. I do not have a good intuition for why the user would be interested in exploring trees based on variable importance.

\textcolor{violet}{The second tab was modifyed in different ways taking into account your valuable comments. We have included three main changes in the second tabset, first we have included jittered points in the OOB error boxplot where the user can select trees based in their performance. Second, we have included a panel to select the nodes to see in the importance variable plot, the mosaic plot and the density plot. The panel allows you to select more or less than three nodes and the mentioned plots are updated acordingly. Finally, we rearrange some of the plots to make the selection and interaction more intuitive  presented in Figure 10.}

@@

 I think the paper overall is good. The comments below are a combination of comments on the paper itself and some suggestions for the plotting in the software (which don't have to be changed for the paper). I think also that some of the details of the plots need to be explained better.

MAIN COMMENTS:

Section 4 in general: since the big point here is that the software is interactive, I think the discussion in this section would be a lot more useful if you illustrated step by step with screenshots an example where your interactive drill-down allows you to quickly discover things. For instance, say there are some outliers in plot A. Then you click on them and see some further details in plot B that help you understand what is going on. In this section you show a bunch of screenshots and label their capabilities, but you don't actually demonstrate an end-to-end example of the interactivity. For instance, in Figure 9 you give an example of a set of actions, why don't you actually show us the results and the changes that occur in a sequence of plots? Furthermore, I would show some actual examples of screenshots of what shows when you hover over observations.

\textcolor{violet}{In Section 4 we have included more details and examples of use and interpreting results to address your comment.}

@@

\item Figures 1 and 2: I see that the potential range for coefficient values for the projection pursuit seem to be between -1 and 1. I imagine this is a feature of the algorithm. Perhaps include in your description of projection pursuit a note about this fact, for those unfamiliar, and mention what optimization criteria is used in it.

\textcolor{violet}{Yes, the coefficient values are between -1 and 1 by construction. We find the linear combinations, \textbf{a} which maximize $\frac{a^TBa}{a^TWa}$, where B y W are the between and within group sums of squares. We get $\hat a_1=e_1$ where $e_1...e_s$ are eigenvectors of $W^{-1}B$. In this case we find the 1-D projection optimizing the LDA index. We have included in Section 3 LDA and PDA indexes which are optimized to get the linear projections and it is more clear now how the coefficients are computed. }

\item Page 6: If you use a notation for VI for global variable importance (you define it as the average across k trees), I'd just define it here as an equation and use it in the text (e.g in line 19).

\textcolor{violet}{Formulas for all VI were defined in equations and mentioned in the text. It  is more clear now.}

@@

\item Page 10, Figure 5: similar remark as before. I think a violin plot or something similar for each class, rather than a jittered dot plot, is more correct. The x-axis for each class is a "fake" axis. I think jittered dot plots should only be used with scatterplots to make sure individual dots can be distinguished.

\textcolor{violet}{We are using points because it is important to see the individual cases and the jitter is to overcome overploting. It is not recommended using jittered scatterplots because you are changing the real values of your data. In this case we are not changing the values of a continuous variable just the position to overcome overplotting in a categorical one.  }

\item Page 10, line 40: This sentence is muddled "As Sievert (2020) describes, one of the biggest difficulties for the app in order manage linking between plots is the data structure management for each widget". Furthermore, what does "widget" refer to here? Use a more descriptive term.

\textcolor{violet}{ At their core Shiny widgets are mini-applications created using the shinyApp function.}

### 2022-06-29 PPforestVizupdate.Rnw
update

<<pkg, echo=FALSE, message=FALSE, warning=FALSE>>=
library(ggplot2)
library(PPforest)
library(shiny)
library(GGally)
library(productplots)
library(PPtreeViz)
library(tidyr)
library(dplyr)
library(pracma)
library(RColorBrewer)
library(randomForest)
library(pROC)
library(gridExtra)
library(grid)
library(ggmosaic)
set.seed(123)
#  ppf2022 <-PPforest(data = crab, class= "Type", std = TRUE, size.tr = 1 , m = 500, PPmethod = 'LDA', size.p =.33,
#   lambda = .1, parallel = FALSE, cores = 2, rule = 1)
# save(ppf2022, file = "ppf2022.RData")
load("ppf2022.RData")
load("rf_fish.RData")
 ppf<- ppf2022
@

>>>>

<<pkg, echo=FALSE, message=FALSE, warning=FALSE>>=
library(ggplot2)
library(PPforest)
library(shiny)
library(GGally)
library(productplots)
library(PPtreeViz)
library(tidyr)
library(dplyr)
library(pracma)
library(RColorBrewer)
library(randomForest)
library(pROC)
library(gridExtra)
library(grid)
library(ggmosaic)
set.seed(123)
#  ppf2022 <-PPforest(data = crab, class= "Type", std = TRUE, size.tr = 1 , m = 500, PPmethod = 'LDA', size.p =.33,
#   lambda = .1, parallel = FALSE, cores = 2, rule = 1)
# save(ppf2022, file = "ppf2022.RData")
load("ppf2022.RData")
load("rf_fish.Rdata")
 ppf<- ppf2022
@
